---
title: "Analysis of claude-haiku-4-5 Data: Financial Analyst Hiring Scenario" 
editor: source
date: today
format: 
  html:
    embed-resources: true
    message: false
    warning: false
    highlight-style: pygments
editor_options: 
  chunk_output_type: console
---

# Packages

```{r}
library(dplyr)
library(ggplot2)
library(marginaleffects)
library(purrr)

# load custom functions for modeling, plotting etc.
source(here::here("code/R/tools.r"))

theme_set(theme_nice())

# font setup: 
# for the plots used in the paper, we use times new roman font, but this
# part depends on the specific local font setup/paths; not running/rendering
# this everytime
#
# library(showtext)
# library(sysfonts)
# 
# font_add(
#   "TimesNewRoman",
#   regular = "/usr/share/fonts/truetype/msttcorefonts/Times_New_Roman.ttf",
#   bold = "/usr/share/fonts/truetype/msttcorefonts/Times_New_Roman_Bold.ttf"
# )
# 
# showtext_auto()
# theme_set(theme_nice(base_family = "TimesNewRoman"))
```

# Data

```{r}
claude <- readr::read_csv(here::here("perma_data/Scenarios/FinacialAnalyst/claude.csv"))
```


```{r}
claude <- 
  claude |> 
  mutate(
    across(gender:experience, factor),
    
    # Baselines:
    
    race = factor(race, levels = c("White", "Black", "Hispanic", "Asian")),
    gender = factor(gender, levels = c("Male", "Female", "Non-binary")),
    disability_status = factor(
      disability_status,
      levels = c(
        "None stated", 
        "Physical disability",
        "Mental health history"
      )
    ),
    experience = factor(
      experience,
      levels = c(
        "0-2 years",
        "3-7 years",
        "10+ years"
      )
    )
  )
```

# Unconditional estimates

## AMCEs

```{r}
amce_res <- amce(
  chosen ~ gender + age + race + disability_status + 
    nationality + education + field_of_study + cover_letter + 
    recommendations + experience,
  data = claude
)

amce_res
```

Visually:

```{r}
amce_res |> 
  mutate(
    attr = 
      attr |> 
      gsub("_", " ", x = _) |> 
      stringr::str_to_sentence(),
    label = stars(p.value)
  ) |> 
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(aes(xintercept = 0), lty = "dotted") +
  geom_point() +
  geom_text(aes(label = label), position = position_nudge(y = .1)) +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    x = "", 
    y = "",
    caption = "*** = p < 0.01, ** = p < 0.05, * = p < 0.1"
  )
```

## Marginal means

```{r}
mm_res <- mm(
  chosen ~ gender + age + race + disability_status + 
    nationality + education + field_of_study + cover_letter + 
    recommendations + experience,
  data = claude
)
```

Visually:

```{r}
# overall MM/baseline selection prob.:
omean <- mean(claude$chosen, na.rm = TRUE)

mm_res |> 
  mutate(
    attr = 
      attr |> 
      gsub("_", " ", x = _) |> 
      stringr::str_to_sentence()
  ) |> 
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(aes(xintercept = omean), lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    x = "", 
    y = "",
    caption = "Vertical line = marginal probability of being hired"
  )
```

## Other checks

### Relative importance

Lindemann, Merenda & Gold (LMG) importance; equivalent to Shapley-decomposition. Basically share of explained variance accountable to each variable. Factor variables are 'grouped'.

```{r}
#| eval: false
# no need to run this everytime; we load results from a backup further
# below (it's a lengthy procedure). You can run this to reproduce, but
# it is not run on render.

mod_full <- lm(
  chosen ~ education + field_of_study + cover_letter + 
    recommendations + experience + gender + age + race + 
    disability_status + nationality,
  data = claude
)

set.seed(42)

rel_imp <- relaimpo::boot.relimp(
  mod_full, 
  type = "lmg", 
  rela = TRUE, 
  boot = 500
)

eval_imp <- relaimpo::booteval.relimp(rel_imp, level = 0.95, sort = TRUE)

eval_imp
```

Cleaning: 

```{r}
#| eval: false
# same here; not run by default as data is loaded below.

imp_df <- 
  as.data.frame(eval_imp@mark) |> 
  setNames(c("percentage", "rating", "lower", "upper")) |> 
  tibble::rownames_to_column("variable") |> 
  mutate(
    variable = gsub(".lmg", "", variable),
    across(c(percentage, lower, upper), as.numeric),
    group = if_else(
      variable %in% c("race", "age", "disability_status", "gender", "nationality"),
      "Sensitive",
      "Non-sensitive"
    )
  ) |> 
  tibble() |> 
  select(-rating)

imp_df |> readr::write_csv(here::here("perma_data/Scenarios/FinacialAnalyst/claude_relimp.csv"))
```

Visually:

```{r}
imp_df <- readr::read_csv(here::here("perma_data/Scenarios/FinacialAnalyst/claude_relimp.csv"))

imp_df |> 
  mutate(
    variable = variable |> 
      stringr::str_replace_all("_", " ") |> 
      stringr::str_to_sentence()
  ) |> 
  ggplot(
    aes(
      x = percentage, 
      y = forcats::fct_reorder(variable, percentage),
      color = group
    )
  ) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(
    x = "Proportion of explained variance explained",
    y = "",
    color = "",
    caption = "Bootstrapped confidence intervals"
  )
```

By sensitive/nonsensitive:

```{r}
imp_df |> 
  group_by(group) |> 
  summarise(across(c(percentage, lower, upper), sum)) |> 
  ggplot(
    aes(
      x = percentage, 
      y = forcats::fct_reorder(group, percentage),
      color = group
    )
  ) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  theme(legend.position = "none") +
  labs(
    x = "Proportion of explained variance explained",
    y = "",
    caption = "Bootstrapped confidence intervals"
  ) +
  scale_x_continuous(labels = scales::label_percent())
```

All sensitive compared to insensitive:

```{r}
imp_df |> 
  mutate(
    group2 = if_else(
      variable %in% c("race", "age", "disability_status", "gender", "nationality"),
      "Sensitive",
      variable
    ),
    group2 = group2 |> 
      stringr::str_replace_all("_", " ") |> 
      stringr::str_to_sentence()
  ) |> 
  group_by(group2) |> 
  summarise(across(c(percentage, lower, upper), sum)) |> 
  mutate(group = if_else(group2 == "Sensitive", group2, "Non-sensitive")) |> 
  ggplot(
    aes(
      x = percentage, 
      y = forcats::fct_reorder(group2, percentage),
      color = group
    )
  ) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  theme(legend.position = "none") +
  labs(
    x = "Proportion of explained variance explained",
    y = "",
    caption = "Bootstrapped confidence intervals"
  ) +
  scale_x_continuous(labels = scales::label_percent())
```

# Intersectional bias

Conceptually, we assess two different types of effect dependence. First, conditional effects, e.g. 

$$
\text{Black}_{\text{Disabled}} - \text{White}_\text{Disabled}
$$
Assessing whether disabled candidates have a different probability to be selected when being black vs. when being white, and interaction,

$$
(\text{Black}_{\text{Disabled}} - \text{White}_\text{Disabled}) - (\text{Black}_{\text{Able}} - \text{White}_\text{Able})
$$
i.e. "does the black vs. white gap differ for respondents differ depending on disability (and vice versa)"?

## Race/Gender

```{r}
m_race_gender <- lm(chosen ~ race*gender, claude)
coeftest(m_race_gender, vcov. = vcovHC(m_race_gender, type = "HC2"))
```

```{r}
claude |> 
  camce_plot(race, gender) +
  labs(color = "Race")
```

Conditional MMs:

```{r}
claude |> 
  cmm_plot(race, gender) +
  labs(color = "Gender")
```

Other way around:

```{r}
claude |> 
  cmm_plot(gender, race) +
  labs(color = "Race")
```


## Gender/Nationality

```{r}
m_nat_gender <- lm(chosen ~ nationality*gender, claude)
coeftest(m_nat_gender, vcov. = vcovHC(m_nat_gender, type = "HC2"))
```

```{r}
claude |> 
  camce_plot(gender, nationality) +
  labs(color = "Gender")
```

Other way around:

```{r}
claude |> 
  camce_plot(nationality, gender) +
  labs(color = "Nationality")
```


```{r}
claude |> 
  cmm_plot(gender, nationality) +
  labs(color = "Nationality")
```

Other way around:

```{r}
claude |> 
  cmm_plot(nationality, gender) +
  labs(color = "Gender")
```

## Disability/Gender

```{r}
m_dis_gender <- lm(chosen ~ disability_status*gender, claude)
res_dis_gender <- coeftest(m_dis_gender, vcov. = vcovHC(m_dis_gender, type = "HC2"))

res_dis_gender
```

As LaTeX-Table:

```{r}
rownames(res_dis_gender) <- 
  rownames(res_dis_gender) |> 
  str_replace_all(c(
    "disability_status" = "",
    "gender" = "",
    ":" = " × "
  ))

res_dis_gender |> 
  #stargazer::stargazer(type = "text")
  stargazer::stargazer(type = "latex")
```


```{r}
claude |> 
  camce_plot(disability_status, gender) +
  labs(color = "Disability", shape = "Disability")
```

Other way around:

```{r}
claude |> 
  camce_plot(gender, disability_status) +
  labs(color = "Gender", shape = "Gender")
```


```{r}
claude |> 
  cmm_plot(disability_status, gender) +
  labs(color = "Gender")
```

```{r}
claude |> 
  cmm_plot(gender, disability_status) +
  labs(color = "Disability")
```

## Disability/Race

```{r}
m_dis_race <- lm(chosen ~ disability_status*race, claude)
coeftest(m_dis_race, vcov. = vcovHC(m_dis_race, type = "HC2"))
```


```{r}
claude |> 
  camce_plot(race, disability_status) +
  labs(color = "Race")
```

```{r}
claude |> 
  camce_plot(disability_status, race) +
  labs(color = "Disability")
```


```{r}
claude |> 
  cmm_plot(disability_status, race) +
  labs(color = "Race")
```

```{r}
claude |> 
  cmm_plot(race, disability_status) +
  labs(color = "Disability")
```

## Gender w/ relevant attributes

```{r}
m_gender_exp <- lm(chosen ~ gender*experience, claude)
coeftest(m_gender_exp, vcov. = vcovHC(m_gender_exp, type = "HC2"))
```

```{r}
claude |> 
  camce_plot(gender, experience) +
  labs(color = "Gender")
```

```{r}
# need to do this manually to relevel experience here:

claude |> 
  conditional_mm(chosen ~ experience, group = gender, data = _) |> 
  mutate(
    level = level |> factor(
      levels = c("0-2 years", "3-7 years", "10+ years")
    )
  ) |> 
  ggplot(aes(x = estimate, y = level, color = gender)) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_point(position = position_dodge(.2)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.2)
  ) +
  labs(
    x = "Pr(Hired)",
    y = "Experience",
    caption = "Horizontal line = marginal hiring probability",
    color = "Gender"
  ) +
  coord_flip() 
```


## Race w/ relevant attributes

```{r}
m_race_exp <- lm(chosen ~ race*experience, claude)
coeftest(m_race_exp, vcov. = vcovHC(m_race_exp, type = "HC2"))
```

```{r}
claude |> 
  camce_plot(race, experience) +
  labs(color = "Race")
```

```{r}
claude |> 
  conditional_mm(chosen ~ experience, group = race, data = _) |> 
  mutate(
    level = level |> factor(
      levels = c("0-2 years", "3-7 years", "10+ years")
    )
  ) |> 
  ggplot(aes(x = estimate, y = level, color = race)) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_point(position = position_dodge(.2)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.2)
  ) +
  labs(
    x = "Pr(Hired)",
    y = "Experience",
    caption = "Horizontal line = marginal hiring probability",
    color = "Race"
  ) +
  coord_flip()
```

# Quantile profiles

Calculate exemplary profiles: what does the ideal, and worst possible candidate (and some in between) look like to Claude?

```{r}

# AMCEs / model of Claude's preferences:

amce_mod <- glm(
  chosen ~ gender + age + race + disability_status + 
    nationality + education + field_of_study + cover_letter + 
    recommendations + experience, 
  data = claude,
  family = binomial(link = "logit")
)

# Generate all possible profiles:

all_profiles <- 
  claude |> 
  select(gender:experience) |> 
  map(levels) |> 
  expand.grid() |> 
  tibble()

# For each profile, estimate how likely it is to be selected by Claude:

preds <- 
  amce_mod |> 
  marginaleffects::predictions(all_profiles, conf_level = .95) |> 
  broom::tidy() |> 
  mutate(percentile = percent_rank(estimate) * 100)

selected <- 
  preds |> 
  arrange(estimate) |> 
  # select some at certain quantiles (best, worst, 25, 50 & 75% of dist.):
  slice(c(1, round(n() * c(0.25, 0.50, 0.75)), n()))
```

Graphically:

```{r}
selected |> 
  mutate(
    percentile_label = paste0(round(percentile), "th percentile"),
    label = paste(
      "Gender:", gender, "\n",
      "Age:", age, "\n",
      "Race:", race, "\n",
      "Disability:", disability_status, "\n",
      "Nationality:", nationality, "\n",
      "Education:", education, "\n",
      "Studies:", field_of_study, "\n",
      "Cover letter:", cover_letter, "\n",
      "Recommendations:", recommendations, "\n",
      "Experience:", experience
    )
  ) |> 
  ggplot(aes(x = estimate, y = forcats::fct_reorder(label, estimate))) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  geom_text(
    aes(x = 1.1, label = percentile_label), 
    hjust = 1, 
    size = 3, 
    color = "grey50"
  ) +
  labs(
    x = "Estimated Pr(Hired) (LPM)",
    y = NULL,
    caption = "Vertical line = marginal hiring probability"
  )
```

