---
title: "Analysis of gpt-5.4-mini Data: Financial Analyst Hiring Scenario"
editor: source
date: today
format: 
  html:
    embed-resources: true
    message: false
    warning: false
    highlight-style: pygments
editor_options: 
  chunk_output_type: console
---

# Packages & Setup

```{r}
library(dplyr)
library(ggplot2)
library(lmtest)
library(marginaleffects)
library(purrr)
library(sandwich)
library(stringr)
library(tidyr)

# load custom functions for modeling, plotting, etc. 
source(here::here("code/R/tools.r"))

theme_set(theme_nice())

# font setup: 
# for the plots used in the paper, we use times new roman font, but this
# part depends on the specific local font setup/paths; not running/rendering
# this everytime
#
# library(showtext)
# library(sysfonts)
# 
# font_add(
#   "TimesNewRoman",
#   regular = "/usr/share/fonts/truetype/msttcorefonts/Times_New_Roman.ttf",
#   bold = "/usr/share/fonts/truetype/msttcorefonts/Times_New_Roman_Bold.ttf"
# )
# 
# showtext_auto()
# theme_set(theme_nice(base_family = "TimesNewRoman"))
```

# Data

```{r}
chatty <- readr::read_csv(here::here("perma_data/Scenarios/FinacialAnalyst/chatgpt.csv"))
```


```{r}
chatty <- 
  chatty |> 
  mutate(
    across(gender:experience, factor),
    
    # Baselines:
    
    race = factor(race, levels = c("White", "Black", "Hispanic", "Asian")),
    gender = factor(gender, levels = c("Male", "Female", "Non-binary")),
    disability_status = factor(
      disability_status,
      levels = c(
        "None stated", 
        "Physical disability",
        "Mental health history"
      )
    ),
    experience = factor(
      experience,
      levels = c(
        "0-2 years",
        "3-7 years",
        "10+ years"
      )
    )
  )
```

# Unconditional estimates

## AMCEs

```{r}
amce_res <- amce(
  chosen ~ gender + age + race + disability_status + 
    nationality + education + field_of_study + cover_letter + 
    recommendations + experience,
  data = chatty
)

amce_res
```

Visually:

```{r}
amce_res |> 
  mutate(
    attr = 
      attr |> 
      gsub("_", " ", x = _) |> 
      stringr::str_to_sentence(),
    label = stars(p.value)
  ) |> 
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(aes(xintercept = 0), lty = "dotted") +
  geom_point(size = 2) +
  geom_text(aes(label = label), position = position_nudge(y = .1)) +
  geom_linerange(aes(xmin = lower, xmax = upper), lwd = 1) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(x = "", y = "")
```

## Marginal means

```{r}
mm_res <-mm(
  chosen ~ gender + age + race + disability_status + 
    nationality + education + field_of_study + cover_letter + 
    recommendations + experience,
  data = chatty
)

mm_res
```

Visually:

```{r}
# overall MM/baseline selection prob.:
omean <- mean(chatty$chosen, na.rm = TRUE)

mm_res |> 
  mutate(
    attr = 
      attr |> 
      gsub("_", " ", x = _) |> 
      stringr::str_to_sentence()
  ) |> 
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(aes(xintercept = omean), lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    x = "", 
    y = "",
    caption = "Vertical line = marginal probability of being hired"
  )
```

## Other checks

### Relative importance

Lindemann, Merenda & Gold (LMG) importance; equivalent to Shapley-decomposition. Basically share of explained variance accountable to each variable. Factor variables are 'grouped'.

```{r}
#| eval: false
# load the bootstrapped results from a backup below, only run this 
# to reproduce (is not re-run on render)

mod_full <- lm(
  chosen ~ education + field_of_study + cover_letter + 
    recommendations + experience + gender + age + race + 
    disability_status + nationality,
  data = chatty
)

set.seed(42)

rel_imp <- relaimpo::boot.relimp(
  mod_full, 
  type = "lmg", 
  rela = TRUE, 
  boot = 500
)

eval_imp <- relaimpo::booteval.relimp(rel_imp, level = 0.95, sort = TRUE)

eval_imp
```

Cleaning the results:

```{r}
#| eval: false
# same here, not run by default since we load results from a backup below

imp_df <- 
  as.data.frame(eval_imp@mark) |> 
  setNames(c("percentage", "rating", "lower", "upper")) |> 
  tibble::rownames_to_column("variable") |> 
  mutate(
    variable = gsub(".lmg", "", variable),
    across(c(percentage, lower, upper), as.numeric),
    group = if_else(
      variable %in% c("race", "age", "disability_status", "gender", "nationality"),
      "Sensitive",
      "Non-sensitive"
    )
  ) |> 
  tibble() |> 
  select(-rating)

# Export so we do not have to re-run this all the time:
imp_df |> readr::write_csv(here::here("perma_data/Scenarios/FinacialAnalyst/chatty_relimp.csv")) 
```

Visually:

```{r}
imp_df <- readr::read_csv(here::here("perma_data/Scenarios/FinacialAnalyst/chatty_relimp.csv"))

imp_df |> 
  mutate(
    variable = variable |> 
      stringr::str_replace_all("_", " ") |> 
      stringr::str_to_sentence()
  ) |> 
  ggplot(
    aes(
      x = percentage, 
      y = forcats::fct_reorder(variable, percentage),
      color = group
    )
  ) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(
    x = "Proportion of explained variance explained",
    y = "",
    color = "",
    caption = "Bootstrapped confidence intervals"
  )
```

By sensitive/nonsensitive:

```{r}
imp_df |> 
  group_by(group) |> 
  summarise(across(c(percentage, lower, upper), sum)) |> 
  ggplot(
    aes(
      x = percentage, 
      y = forcats::fct_reorder(group, percentage),
      color = group
    )
  ) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  theme(legend.position = "none") +
  labs(
    x = "Proportion of explained variance explained",
    y = "",
    caption = "Bootstrapped confidence intervals"
  ) +
  scale_x_continuous(labels = scales::label_percent())
```

# Intersectional bias

Conceptually, assess two different types of effect dependence. First, conditional effects, e.g. 

$$
\text{Black}_{\text{Disabled}} - \text{White}_\text{Disabled}
$$
Assessing whether disabled candidates have a different probability to be selected when being black vs. when being white, and interaction,

$$
(\text{Black}_{\text{Disabled}} - \text{White}_\text{Disabled}) - (\text{Black}_{\text{Able}} - \text{White}_\text{Able})
$$
i.e. "does the black vs. white gap differ for respondents differ depending on disability (and vice versa)"?

## Race/Gender

Look at conditional marginal means: for whites (baseline) and blacks, non-binary candidates are less likely to be picked than males or females. For hispanic and asian candidates, however, non-binary candidates are not penalized; gender operates differently across races.

```{r}
m_race_gender <- lm(chosen ~ race*gender, chatty)
coeftest(m_race_gender, vcov. = vcovHC(m_race_gender, type = "HC2"))
```

Conditional AMCEs. Among non-binary and female candidates, the model discriminates against black applicants: black non-binaries and black women are significantly less likely to be selected than their white counterparts.

```{r}
chatty |> 
  camce_plot(race, gender) +
  labs(color = "Race")
```

Conditional MMs:

```{r}

chatty |> 
  cmm_plot(gender, race) +
  labs(color = "race")

# As Heatmap
#
# cmm_gender_race |> 
#   ggplot(aes(x = level, y = race, fill = estimate)) +
#   geom_tile() +
#   scale_fill_distiller(
#     palette = "RdBu",
#     direction = -1,
#     labels = scales::label_percent(suffix = "pp.")
#   ) +
#   theme_minimal()
```

## Gender/Nationality

Gendered effects also differ heavily across nationality. Look at CMMs: while for Indian, German, and American applicants women are generally preferred (significant at 5% for Indians and Americans, but not Germans), the dynamic is reversed for Mexicans and Canadians (also significant interaction terms). However, CAMCEs indicate that women are not significantly discriminated against (Canada p > 0.1, Mexico 0.1 > p > 0.05).

```{r}
m_nat_gender <- lm(chosen ~ nationality*gender, chatty)
coeftest(m_nat_gender, vcov. = vcovHC(m_nat_gender, type = "HC2"))
```


```{r}
chatty |> 
  camce_plot(gender, nationality) +
  labs(color = "Gender")
```

```{r}
chatty |> 
  cmm_plot(nationality, gender) +
  labs(color = "Gender")
```

## Disability/Gender

Effect dynamics of disability are similar across genders (no significant interaction terms, also look at CMMs). However, among candidates with a mental health history, women are significantly more likely to be selected for hiring than men by ChatGPT.

```{r}
m_dis_gender <- lm(chosen ~ disability_status*gender, chatty)

res_dis_gender <- coeftest(
  m_dis_gender, 
  vcov. = vcovHC(m_dis_gender, type = "HC2")
)

res_dis_gender
```

Nicely format this into a regression table, latex output:

```{r}
rownames(res_dis_gender) <- 
  rownames(res_dis_gender) |> 
  str_replace_all(c(
    "disability_status" = "",
    "gender" = "",
    ":" = " × "
  ))

res_dis_gender |> 
  #stargazer::stargazer(type = "text")
  stargazer::stargazer(type = "latex")
```


Females significantly more likely to still be hired than males when both have a mental health history:

```{r}
chatty |> 
  camce_plot(gender, disability_status) +
  theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  labs(color = "Gender", shape = "Gender") +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.9, 0.7),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 0)
  ) +
  scale_y_discrete(
    labels = (\(x, ...) gsub("\\s", "\n", x))(levels(chatty$disability_status))
    #expand = expansion(add = c(1, 1))
  ) +
  xlim(c(-.04, .04))


# ggsave(
#   here::here("writing/tex/figures/gender_disability_chatty.pdf"),
#   width = 5,
#   height = 3
# )
```

Physical disability affects all genders more or less evenly, but women are most likely to be hired compared to other genders both in the absence of a disability, and when there is a mental health history.

```{r}
chatty |> 
  cmm_plot(disability_status, gender) +
  #theme_nice(base_size = 14, base_family = "TimesNewRoman") +
  labs(color = "Gender", shape = "Gender") +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.9, 0.85),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 0)
  ) +
  scale_y_discrete(
    labels = (\(x, ...) gsub("\\s", "\n", x))(levels(chatty$disability_status))
  ) +
  xlim(c(omean - .06, omean + .06))

# ggsave(
#   here::here("writing/tex/figures/gender_disability_chatty_mm.pdf"),
#   width = 5,
#   height = 4
# )
```

## Disability/Race

The impact of disability on hiring decisions made by ChatGPT also varies across races. Looking at interaction terms and the CMMs, we can see that disability operates differently for Hispanic candidates: the premium for being able-bodied is much weaker for Hispanics.

```{r}
m_race_dis <- lm(chosen ~ race*disability_status, chatty)
coeftest(m_race_dis, vcov. = vcovHC(m_race_dis, type = "HC2"))
```

Looking at conditional AMCEs, among candidates with a mental health history, blacks are significantly less likely to be hired than whites.

```{r}
chatty |> 
  camce_plot(race, disability_status) +
  labs(color = "Race")
```

When comparing marginal means, we can see the general dynamic of blacks being penalized additionally for being disabled.

```{r}
chatty |> 
  cmm_plot(race, disability_status) +
  labs(color = "Race")
```


```{r}
chatty |> 
  cmm_plot(disability_status, race) +
  labs(color = "Disability")
```


## Gender w/ relevant attributes

The effect of relevant attributes, like experience, also differs across genders. 

e can see that men are favored at high levels of professional experience. At low to moderate levels of experience, there is generally a preference for women. However, at high levels of experience (10+ years), this effect reverses, with Chatty favoring men.

```{r}
m_gender_exp <- lm(chosen ~ gender*experience, chatty)
coeftest(m_gender_exp, vcov. = vcovHC(m_gender_exp, type = "HC2"))
```

```{r}
chatty |> 
  camce_plot(gender, experience) +
  labs(color = "Gender")
```


```{r}
# need to do this manually to relevel experience here:

chatty |> 
  conditional_mm(chosen ~ experience, group = gender, data = _) |> 
  mutate(
    level = level |> factor(
      levels = c("0-2 years", "3-7 years", "10+ years")
    )
  ) |> 
  ggplot(aes(x = estimate, y = level, color = gender)) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_point(position = position_dodge(.2)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.2)
  ) +
  labs(
    x = "Pr(Hired)",
    y = "Experience",
    caption = "Horizontal line = marginal hiring probability",
    color = "Gender"
  ) +
  coord_flip() 
```

## Race w/ relevant attributes

Experience is valued differently according to race. While black respondents are generally less preferred than Whites by ChatGPT, experience is also less valued (significant interaction term meaning effect of high experience is significantly weaker for Blacks). 

CAMCEs: among high-experience candidates, Blacks significantly lower probability of being hired by ChatGPT than Whites.

CMMs: Blacks generally underselected at all experience levels, but difference greatest among most experienced.

```{r}
m_race_exp <- lm(chosen ~ race*experience, chatty)
res_race_exp <- coeftest(m_race_exp, vcov. = vcovHC(m_race_exp, type = "HC2"))

res_race_exp
```

Regression table:

```{r}
rownames(res_race_exp) <- 
  rownames(res_race_exp) |> 
  str_replace_all(c(
    "experience" = "",
    "race" = "",
    ":" = " × "
  ))

res_race_exp |> 
  #stargazer::stargazer(type = "text")
  stargazer::stargazer(type = "latex")
```


```{r}
chatty |> 
  conditional_amce(chosen ~ race, group = experience, data = _) |> 
  mutate(label = stars(p.value)) |> 
  ggplot(aes(x = experience, y = estimate, color = level, shape = level)) +
  geom_hline(yintercept = 0, lty = "dotted") +
  geom_point(position = position_dodge(.45)) +
  geom_linerange(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(.45)
  ) +
  geom_text(
    aes(label = label), 
    position = position_dodge(.45),
    hjust = 1.1,
    vjust = .1, 
    show.legend = FALSE
  ) +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  #theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.2, 0.2)
  ) +
  labs(
    x = "Change in Pr(Hired)",
    y = "",
    color = "Race",
    shape = "Race"
  )

# ggsave(
#   here::here("writing/tex/figures/race_experience_chatty.pdf"),
#   width = 7, 
#   height = 5
# )
```

```{r}
chatty |> 
  conditional_amce(chosen ~ race, group = experience, data = _) |> 
  filter(level != "White") |> 
  mutate(label = stars(p.value)) |> 
  ggplot(aes(x = experience, y = estimate, color = level, shape = level)) +
  geom_hline(yintercept = 0, lty = "dotted") +
  geom_ribbon(
    aes(ymin = lower, ymax = upper, group = level, fill = level),
    alpha = .1,
    color = NA
  ) +
  geom_point() +
  geom_line(aes(group = level)) +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  rcartocolor::scale_fill_carto_d(palette = "Prism") +
  #theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.3, 0.2)
  ) +
  labs(
    x = "Experience",
    y = "Conditional AMCE (vs. White)",
    color = "Race",
    shape = "Race",
    fill = "Race"
  )

# ggsave(
#   here::here("writing/tex/figures/race_experience_chatty_alt.pdf"),
#   width = 7, 
#   height = 5
# )
```


Conditional marginal means:

```{r}
chatty |> 
  conditional_mm(chosen ~ experience, group = race, data = _) |> 
  mutate(
    level = level |> factor(
      levels = c("0-2 years", "3-7 years", "10+ years")
    )
  ) |> 
  ggplot(aes(x = estimate, y = level, color = race, shape = race)) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_point(position = position_dodge(.2)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.2)
  ) +
  labs(
    x = "Pr(Hired)",
    y = "Experience",
    color = "Race",
    shape = "Race"
  ) +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  coord_flip() +
  #theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  theme(
    axis.line.x = element_line(color = "grey50"),
    axis.line.y = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_line(color = "grey50"),
    legend.position = "inside", 
    legend.position.inside = c(0.9, 0.25)
  )

# ggsave(
#   here::here("writing/tex/figures/race_experience_chatty_cmm.pdf"),
#   width = 7,
#   height = 5
# )
```




# Quantile profiles

Calculate exemplary profiles: what does the ideal, and worst possible candidate (and some in between) look like to ChatGPT?

```{r}

# AMCEs / model of chatty's preferences:

amce_mod <- 
  glm(
    chosen ~ gender + age + race + disability_status + 
      nationality + education + field_of_study + cover_letter + 
      recommendations + experience, 
    data = chatty,
    family = binomial(link = "logit")
  )

# Generate all possible profiles:

all_profiles <- 
  chatty |> 
  select(gender:experience) |> 
  map(levels) |> 
  expand.grid() |> 
  tibble()

# For each profile, estimate how likely it is to be selected by chatty:

preds <- 
  amce_mod |> 
  marginaleffects::predictions(all_profiles, conf_level = .95) |> 
  broom::tidy() |> 
  mutate(percentile = percent_rank(estimate) * 100)

selected <- 
  preds |> 
  arrange(estimate) |> 
  # select some at certain quantiles (best, worst, 25, 50 & 75% of dist.):
  slice(c(1, round(n() * c(0.25, 0.50, 0.75)), n()))
```

Graphically:

```{r}

selected |> 
  mutate(
    percentile_label = paste0(round(percentile), "th percentile"),
    label = paste(
      "Gender:", gender, "\n",
      "Age:", age, "\n",
      "Race:", race, "\n",
      "Disability:", disability_status, "\n",
      "Nationality:", nationality, "\n",
      "Education:", education, "\n",
      "Studies:", field_of_study, "\n",
      "Cover letter:", cover_letter, "\n",
      "Recommendations:", recommendations, "\n",
      "Experience:", experience
    )
  ) |> 
  ggplot(aes(x = estimate, y = forcats::fct_reorder(label, estimate))) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  geom_text(
    aes(x = 1.1, label = percentile_label), 
    hjust = 1, 
    size = 3, 
    color = "grey50"
  ) +
  labs(
    x = "Estimated Pr(Hired) (LPM)",
    y = NULL,
    caption = "Vertical line = marginal hiring probability"
  )
```

