---
title: "Robustness Checks"
editor: source
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    message: false
    warning: false
    highlight-style: pygments
editor_options:
  chunk_output_type: console
---

```{r}
library(dplyr)
library(ggplot2)
library(lmtest)
library(purrr)
library(sandwich)
library(stringr)
library(tidyr)

source(here::here("code/R/tools.r"))

theme_set(theme_nice())
```

# Data

```{r}
chatty <- readr::read_csv(
  here::here("perma_data/Scenarios/FinacialAnalyst/chatgpt.csv"),
  show_col_types = FALSE
) |> filter(!is.na(chosen))

claude <- readr::read_csv(
  here::here("perma_data/Scenarios/FinacialAnalyst/claude.csv"),
  show_col_types = FALSE
) |> filter(!is.na(chosen))
```

## Observation counts

```{r}
tibble(
  Model             = c("GPT-5.4 mini", "Claude Haiku 4.5"),
  `Valid obs`       = c(nrow(chatty), nrow(claude)),
  Selected          = c(sum(chatty$chosen), sum(claude$chosen)),
  Rejected          = c(sum(chatty$chosen == 0), sum(claude$chosen == 0)),
  `Selection rate`  = scales::percent(
    c(mean(chatty$chosen), mean(claude$chosen)),
    accuracy = 0.1
  )
) |>
  knitr::kable()
```

# Construct validity: known-groups design

I run GPT-5.4 mini under three conditions and estimated AMCEs for gender and race.

```{r}
prep <- function(df) {
  df |>
    mutate(
      across(gender:experience, factor),
      race = factor(race, levels = c("White", "Black", "Hispanic", "Asian")),
      gender = factor(gender, levels = c("Male", "Female", "Non-binary")),
      disability_status = factor(
        disability_status,
        levels = c("None stated", "Physical disability", "Mental health history")
      ),
      experience = factor(
        experience,
        levels = c("0-2 years", "3-7 years", "10+ years")
      )
    )
}

# neutral <- readr::read_csv(
#   here::here("perma_data/MeasurmentValidity/Sanity_neutralPrompt.csv"),
#   show_col_types = FALSE
# ) |> prep()

neutral <- prep(chatty) # just use the normal data, was also build with neutral prompt

anti <- readr::read_csv(
  here::here("perma_data/MeasurmentValidity/Sanity_AntiDiversity_FinancialAnalyst_long.csv"),
  show_col_types = FALSE
) |> prep()

pro <- readr::read_csv(
  here::here("perma_data/MeasurmentValidity/Sanity_ProDiversity_FinancialAnalyst_long.csv"),
  show_col_types = FALSE
) |> prep()
```

```{r}
tibble(
  Condition     = c("Anti-diversity", "Neutral", "Pro-diversity"),
  `N obs`       = c(nrow(anti), nrow(neutral), nrow(pro)),
  `Sel. rate`   = scales::percent(
    c(mean(anti$chosen), mean(neutral$chosen), mean(pro$chosen)),
    accuracy = 0.1
  )
) |>
  knitr::kable()
```

## AMCEs: gender and race

```{r}
full_fml <- chosen ~ gender + age + race + disability_status + nationality +
  education + field_of_study + cover_letter + recommendations + experience

get_amce_gender_race <- function(df, label) {
  m   <- lm(full_fml, data = df)
  res <- coeftest(m, vcov. = vcovHC(m, type = "HC2"))
  broom::tidy(res) |>
    filter(grepl("^genderFemale|^raceBlack|^raceHispanic|^raceAsian", term)) |>
    mutate(
      Comparison = case_when(
        term == "genderFemale"  ~ "Male vs. Female",
        term == "raceBlack"     ~ "Black vs. White",
        term == "raceHispanic"  ~ "Hispanic vs. White",
        term == "raceAsian"     ~ "Asian vs. White"
      ),
      # Male vs. Female = negative of the Female AMCE; race comparisons are direct
      `AMCE (pp)` = round(
        ifelse(term == "genderFemale", -estimate, estimate) * 100,
        1
      ),
      sig = stars(p.value),
      condition = label
    ) |>
    select(condition, Comparison, `AMCE (pp)`, sig, p.value)
}

validity_res <-
  bind_rows(
    get_amce_gender_race(anti,    "Anti-diversity"),
    get_amce_gender_race(neutral, "Neutral"),
    get_amce_gender_race(pro,     "Pro-diversity")
  )

validity_res |>
  mutate(`AMCE (pp)` = paste0(`AMCE (pp)`, sig)) |>
  select(-sig, -p.value) |>
  pivot_wider(names_from = condition, values_from = `AMCE (pp)`) |>
  knitr::kable(
    caption = "AMCEs (pp) for gender and race under three prompt conditions. * p<0.1, ** p<0.05, *** p<0.01.",
    format = "latex"
  )
```

Visually:

```{r}
validity_res |>
  mutate(
    condition = factor(
      condition,
      levels = c("Anti-diversity", "Neutral", "Pro-diversity")
    ),
    Comparison = factor(
      Comparison,
      levels = c("Male vs. Female", "Black vs. White", "Hispanic vs. White", "Asian vs. White")
    )
  ) |>
  ggplot(aes(x = `AMCE (pp)`, y = Comparison, color = condition, shape = condition)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point(position = position_dodge(width = 0.4), size = 2) +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  labs(
    x = "AMCE (percentage points)",
    y = "",
    color = "Condition",
    shape = "Condition",
    caption = "Full model with all attributes; HC2-robust SEs."
  ) +
  theme(legend.position = "bottom")
```

# Measurement reliability: decision stability

## Verbatim reruns (Claude Haiku 4.5)

The same 10,000 profiles were submitted three times with identical prompts.

```{r}
c1 <- readr::read_csv(
  here::here("perma_data/Scenarios/FinacialAnalyst/other_data/claude_data/Claude_1.csv"),
  show_col_types = FALSE
)
c2 <- readr::read_csv(
  here::here("perma_data/Scenarios/FinacialAnalyst/other_data/claude_data/Claude_2.csv"),
  show_col_types = FALSE
)
c3 <- readr::read_delim(
  here::here("perma_data/Scenarios/FinacialAnalyst/other_data/claude_data/Claude_3.csv"),
  delim = ";",
  show_col_types = FALSE
)

n   <- nrow(c1)
all_agree  <- (c1$chosen == c2$chosen) & (c2$chosen == c3$chosen)
n_agree    <- sum(all_agree, na.rm = TRUE)
n_disagree <- n - n_agree

tibble(
  `N profiles`          = n,
  `All 3 agree`         = n_agree,
  `At least one differs` = n_disagree,
  `Agreement rate`      = scales::percent(n_agree / n, accuracy = 0.1)
) |>
  knitr::kable()
```

Pairwise breakdown:

```{r}
tibble(
  Pair = c("Run 1 vs. Run 2", "Run 1 vs. Run 3", "Run 2 vs. Run 3"),
  Agreements = c(
    sum(c1$chosen == c2$chosen, na.rm = TRUE),
    sum(c1$chosen == c3$chosen, na.rm = TRUE),
    sum(c2$chosen == c3$chosen, na.rm = TRUE)
  )
) |>
  mutate(
    Disagreements = n - Agreements,
    Rate = scales::percent(Agreements / n, accuracy = 0.1)
  ) |>
  knitr::kable()
```

## Within-dataset order-varied repeats (Claude Haiku 4.5)

Some profiles in the main dataset were drawn more than once (with attribute order
re-randomized each time).  Agreement rate on these repeated profiles measures
sensitivity to prompt order.

```{r}
profile_cols <- c(
  "gender", "age", "race", "disability_status", "nationality",
  "education", "field_of_study", "cover_letter", "recommendations", "experience"
)

dup_analysis <- claude |>
  group_by(across(all_of(profile_cols))) |>
  summarise(n = n(), n_distinct_choices = n_distinct(chosen), .groups = "drop")

repeated  <- dup_analysis |> filter(n > 1)
n_agree   <- sum(repeated$n_distinct_choices == 1)

tibble(
  `Unique profile combos`          = nrow(dup_analysis),
  `Seen once`                      = sum(dup_analysis$n == 1),
  `Seen 2+ times`                  = nrow(repeated),
  `Total obs in repeated profiles` = sum(repeated$n),
  `Full agreement`                 = n_agree,
  `Agreement rate`                 = scales::percent(n_agree / nrow(repeated), accuracy = 0.1)
) |>
  knitr::kable()
```
