---
title: "Analysis of Mistral-7B-Instruct-v0.2 Data: Financial Analyst Hiring Scenario"
editor: source
date: today
format: 
  html:
    embed-resources: true
    message: false
    warning: false
    highlight-style: pygments
editor_options: 
  chunk_output_type: console
---

# Packages & Setup

```{r}
library(dplyr)
library(ggplot2)
library(lmtest)
library(marginaleffects)
library(purrr)
library(sandwich)
library(stringr)
library(tidyr)

# load custom functions for modeling, plotting, etc. 
source(here::here("code/R/tools.r"))

theme_set(theme_nice())

# font setup: 
# for the plots used in the paper, we use times new roman font, but this
# part depends on the specific local font setup/paths; not running/rendering
# this everytime
#
# library(showtext)
# library(sysfonts)
# 
# font_add(
#   "TimesNewRoman",
#   regular = "/usr/share/fonts/truetype/msttcorefonts/Times_New_Roman.ttf",
#   bold = "/usr/share/fonts/truetype/msttcorefonts/Times_New_Roman_Bold.ttf"
# )
# 
# showtext_auto()
# theme_set(theme_nice(base_family = "TimesNewRoman"))
```

# Data

```{r}
mistral <- readr::read_csv(here::here("perma_data/Scenarios/FinacialAnalyst/mistral.csv"))
names(mistral)
```


```{r}
mistral <- 
  mistral |> 
  mutate(
    across(gender:experience, factor),
    
    # Baselines:
    
    race = factor(race, levels = c("White", "Black", "Hispanic", "Asian")),
    gender = factor(gender, levels = c("Male", "Female", "Non-binary")),
    disability_status = factor(
      disability_status,
      levels = c(
        "None stated", 
        "Physical disability",
        "Mental health history"
      )
    ),
    experience = factor(
      experience,
      levels = c(
        "0-2 years",
        "3-7 years",
        "10+ years"
      )
    )
  )
```

# Unconditional estimates

## AMCEs

```{r}
amce_res <- amce(
  chosen ~ gender + age + race + disability_status + 
    nationality + education + field_of_study + cover_letter + 
    recommendations + experience,
  data = mistral
)

amce_res
```

Visually:

```{r}
amce_res |> 
  mutate(
    attr = 
      attr |> 
      gsub("_", " ", x = _) |> 
      stringr::str_to_sentence(),
    label = stars(p.value)
  ) |> 
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(aes(xintercept = 0), lty = "dotted") +
  geom_point(size = 2) +
  geom_text(aes(label = label), position = position_nudge(y = .1)) +
  geom_linerange(aes(xmin = lower, xmax = upper), lwd = 1) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(x = "", y = "")
```

## Marginal means

```{r}
mm_res <-mm(
  chosen ~ gender + age + race + disability_status + 
    nationality + education + field_of_study + cover_letter + 
    recommendations + experience,
  data = mistral
)

mm_res
```

Visually:

```{r}
# overall MM/baseline selection prob.:
omean <- mean(mistral$chosen, na.rm = TRUE)

mm_res |> 
  mutate(
    attr = 
      attr |> 
      gsub("_", " ", x = _) |> 
      stringr::str_to_sentence()
  ) |> 
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(aes(xintercept = omean), lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    x = "", 
    y = "",
    caption = "Vertical line = marginal probability of being hired"
  )
```

## Other checks

### Relative importance

Lindemann, Merenda & Gold (LMG) importance; equivalent to Shapley-decomposition. Basically share of explained variance accountable to each variable. Factor variables are 'grouped'.

```{r}
# runs bootstrap only if cached CSV does not exist yet; otherwise loads from file
relimp_path <- here::here("perma_data/Scenarios/FinacialAnalyst/mistral_relimp.csv")

if (!file.exists(relimp_path)) {
  mod_full <- lm(
    chosen ~ education + field_of_study + cover_letter +
      recommendations + experience + gender + age + race +
      disability_status + nationality,
    data = mistral
  )

  set.seed(42)

  rel_imp <- relaimpo::boot.relimp(
    mod_full,
    type = "lmg",
    rela = TRUE,
    boot = 500
  )

  eval_imp <- relaimpo::booteval.relimp(rel_imp, level = 0.95, sort = TRUE)

  imp_df <-
    as.data.frame(eval_imp@mark) |>
    setNames(c("percentage", "rating", "lower", "upper")) |>
    tibble::rownames_to_column("variable") |>
    mutate(
      variable = gsub(".lmg", "", variable),
      across(c(percentage, lower, upper), as.numeric),
      group = if_else(
        variable %in% c("race", "age", "disability_status", "gender", "nationality"),
        "Sensitive",
        "Non-sensitive"
      )
    ) |>
    tibble() |>
    select(-rating)

  imp_df |> readr::write_csv(relimp_path)
} else {
  imp_df <- readr::read_csv(relimp_path)
}
```

Visually:

```{r}
imp_df |>
  mutate(
    variable = variable |> 
      stringr::str_replace_all("_", " ") |> 
      stringr::str_to_sentence()
  ) |> 
  ggplot(
    aes(
      x = percentage, 
      y = forcats::fct_reorder(variable, percentage),
      color = group
    )
  ) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(
    x = "Proportion of explained variance explained",
    y = "",
    color = "",
    caption = "Bootstrapped confidence intervals"
  )
```

By sensitive/nonsensitive:

```{r}
imp_df |> 
  group_by(group) |> 
  summarise(across(c(percentage, lower, upper), sum)) |> 
  ggplot(
    aes(
      x = percentage, 
      y = forcats::fct_reorder(group, percentage),
      color = group
    )
  ) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  theme(legend.position = "none") +
  labs(
    x = "Proportion of explained variance explained",
    y = "",
    caption = "Bootstrapped confidence intervals"
  ) +
  scale_x_continuous(labels = scales::label_percent())
```

# Intersectional bias

Conceptually, we assess two different types of effect dependence. First, conditional effects, e.g. 

$$
\text{Black}_{\text{Disabled}} - \text{White}_\text{Disabled}
$$
Assessing whether disabled candidates have a different probability to be selected when being black vs. when being white, and interaction,

$$
(\text{Black}_{\text{Disabled}} - \text{White}_\text{Disabled}) - (\text{Black}_{\text{Able}} - \text{White}_\text{Able})
$$
i.e. "does the black vs. white gap differ for respondents differ depending on disability (and vice versa)"?

## Race/Gender

```{r}
m_race_gender <- lm(chosen ~ race*gender, mistral)
coeftest(m_race_gender, vcov. = vcovHC(m_race_gender, type = "HC2"))
```

Conditional AMCEs:

```{r}
mistral |>
  camce_plot(race, gender) +
  labs(color = "Race")
```

Conditional MMs:

```{r}

mistral |> 
  cmm_plot(gender, race) +
  labs(color = "race")

# Heatmap
#
# cmm_gender_race |> 
#   ggplot(aes(x = level, y = race, fill = estimate)) +
#   geom_tile() +
#   scale_fill_distiller(
#     palette = "RdBu",
#     direction = -1,
#     labels = scales::label_percent(suffix = "pp.")
#   ) +
#   theme_minimal()
```

## Gender/Nationality

```{r}
m_nat_gender <- lm(chosen ~ nationality*gender, mistral)
coeftest(m_nat_gender, vcov. = vcovHC(m_nat_gender, type = "HC2"))
```


```{r}
mistral |> 
  camce_plot(gender, nationality) +
  labs(color = "Gender")
```

```{r}
mistral |> 
  cmm_plot(nationality, gender) +
  labs(color = "Gender")
```

## Disability/Gender

```{r}
m_dis_gender <- lm(chosen ~ disability_status*gender, mistral)

res_dis_gender <- coeftest(
  m_dis_gender, 
  vcov. = vcovHC(m_dis_gender, type = "HC2")
)

res_dis_gender
```

Nicely format this into a regression table, latex output:

```{r}
rownames(res_dis_gender) <- 
  rownames(res_dis_gender) |> 
  str_replace_all(c(
    "disability_status" = "",
    "gender" = "",
    ":" = " × "
  ))

res_dis_gender |> 
  #stargazer::stargazer(type = "text")
  stargazer::stargazer(type = "latex")
```


Conditional AMCEs:

```{r}
mistral |>
  camce_plot(gender, disability_status) +
  theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  labs(color = "Gender", shape = "Gender") +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.9, 0.7),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 0)
  ) +
  scale_y_discrete(
    labels = (\(x, ...) gsub("\\s", "\n", x))(levels(mistral$disability_status))
    #expand = expansion(add = c(1, 1))
  ) +
  xlim(c(-.04, .04))


# ggsave(
#   here::here("writing/tex/figures/gender_disability_mistral.pdf"),
#   width = 5,
#   height = 3
# )
```

Conditional MMs:

```{r}
mistral |>
  cmm_plot(disability_status, gender) +
  #theme_nice(base_size = 14, base_family = "TimesNewRoman") +
  labs(color = "Gender", shape = "Gender") +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.9, 0.85),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 0)
  ) +
  scale_y_discrete(
    labels = (\(x, ...) gsub("\\s", "\n", x))(levels(mistral$disability_status))
  ) +
  xlim(c(omean - .06, omean + .06))

# ggsave(
#   here::here("writing/tex/figures/gender_disability_mistral_mm.pdf"),
#   width = 5,
#   height = 4
# )
```

## Disability/Race

```{r}
m_race_dis <- lm(chosen ~ race*disability_status, mistral)
coeftest(m_race_dis, vcov. = vcovHC(m_race_dis, type = "HC2"))
```

Conditional AMCEs:

```{r}
mistral |>
  camce_plot(race, disability_status) +
  labs(color = "Race")
```

Conditional MMs:

```{r}
mistral |>
  cmm_plot(race, disability_status) +
  labs(color = "Race")
```

```{r}
mistral |>
  cmm_plot(disability_status, race) +
  labs(color = "Disability")
```


## Gender w/ relevant attributes

The effect of relevant attributes, like experience, also differs across genders.

```{r}
m_gender_exp <- lm(chosen ~ gender*experience, mistral)
coeftest(m_gender_exp, vcov. = vcovHC(m_gender_exp, type = "HC2"))
```

```{r}
mistral |> 
  camce_plot(gender, experience) +
  labs(color = "Gender")
```


```{r}
# need to do this manually to relevel experience here:

mistral |> 
  conditional_mm(chosen ~ experience, group = gender, data = _) |> 
  mutate(
    level = level |> factor(
      levels = c("0-2 years", "3-7 years", "10+ years")
    )
  ) |> 
  ggplot(aes(x = estimate, y = level, color = gender)) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_point(position = position_dodge(.2)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.2)
  ) +
  labs(
    x = "Pr(Hired)",
    y = "Experience",
    caption = "Horizontal line = marginal hiring probability",
    color = "Gender"
  ) +
  coord_flip() # flipping (I think experience makes more sense this way)
```

## Race w/ relevant attributes

```{r}
m_race_exp <- lm(chosen ~ race*experience, mistral)
res_race_exp <- coeftest(m_race_exp, vcov. = vcovHC(m_race_exp, type = "HC2"))

res_race_exp
```

Regression table:

```{r}
rownames(res_race_exp) <- 
  rownames(res_race_exp) |> 
  str_replace_all(c(
    "experience" = "",
    "race" = "",
    ":" = " × "
  ))

res_race_exp |> 
  #stargazer::stargazer(type = "text")
  stargazer::stargazer(type = "latex")
```


```{r}
mistral |> 
  conditional_amce(chosen ~ race, group = experience, data = _) |> 
  mutate(label = stars(p.value)) |> 
  ggplot(aes(x = experience, y = estimate, color = level, shape = level)) +
  geom_hline(yintercept = 0, lty = "dotted") +
  geom_point(position = position_dodge(.45)) +
  geom_linerange(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(.45)
  ) +
  geom_text(
    aes(label = label), 
    position = position_dodge(.45),
    hjust = 1.1,
    vjust = .1, 
    show.legend = FALSE
  ) +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  #theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.2, 0.2)
  ) +
  labs(
    x = "Change in Pr(Hired)",
    y = "",
    color = "Race",
    shape = "Race"
  )

# ggsave(
#   here::here("writing/tex/figures/race_experience_mistral.pdf"),
#   width = 7, 
#   height = 5
# )
```

```{r}
mistral |> 
  conditional_amce(chosen ~ race, group = experience, data = _) |> 
  filter(level != "White") |> 
  mutate(label = stars(p.value)) |> 
  ggplot(aes(x = experience, y = estimate, color = level, shape = level)) +
  geom_hline(yintercept = 0, lty = "dotted") +
  geom_ribbon(
    aes(ymin = lower, ymax = upper, group = level, fill = level),
    alpha = .1,
    color = NA
  ) +
  geom_point() +
  geom_line(aes(group = level)) +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  rcartocolor::scale_fill_carto_d(palette = "Prism") +
  #theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  theme(
    legend.position = "inside", 
    legend.position.inside = c(0.3, 0.2)
  ) +
  labs(
    x = "Experience",
    y = "Conditional AMCE (vs. White)",
    color = "Race",
    shape = "Race",
    fill = "Race"
  )

# ggsave(
#   here::here("writing/tex/figures/race_experience_mistral_alt.pdf"),
#   width = 7, 
#   height = 5
# )
```


Conditional marginal means:

```{r}
# Nice for paper:

mistral |> 
  conditional_mm(chosen ~ experience, group = race, data = _) |> 
  mutate(
    level = level |> factor(
      levels = c("0-2 years", "3-7 years", "10+ years")
    )
  ) |> 
  ggplot(aes(x = estimate, y = level, color = race, shape = race)) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_point(position = position_dodge(.2)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.2)
  ) +
  labs(
    x = "Pr(Hired)",
    y = "Experience",
    color = "Race",
    shape = "Race"
  ) +
  rcartocolor::scale_color_carto_d(palette = "Prism") +
  coord_flip() +
  #theme_nice(base_size = 16, base_family = "TimesNewRoman") +
  theme(
    axis.line.x = element_line(color = "grey50"),
    axis.line.y = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_line(color = "grey50"),
    legend.position = "inside", 
    legend.position.inside = c(0.9, 0.25)
  )

# ggsave(
#   here::here("writing/tex/figures/race_experience_mistral_cmm.pdf"),
#   width = 7,
#   height = 5
# )
```



# Quantile profiles

Calculate exemplary profiles: what does the ideal, and worst possible candidate (and some in between) look like to Mistral?

```{r}

# AMCEs / model of mistral's preferences:

amce_mod <- 
  glm(
    chosen ~ gender + age + race + disability_status + 
      nationality + education + field_of_study + cover_letter + 
      recommendations + experience, 
    data = mistral,
    family = binomial(link = "logit")
  )

# Generate all possible profiles:

all_profiles <- 
  mistral |> 
  select(gender:experience) |> 
  map(levels) |> 
  expand.grid() |> 
  tibble()

# For each profile, estimate how likely it is to be selected by mistral:

preds <- 
  amce_mod |> 
  marginaleffects::predictions(all_profiles, conf_level = .95) |> 
  broom::tidy() |> 
  mutate(percentile = percent_rank(estimate) * 100)

selected <- 
  preds |> 
  arrange(estimate) |> 
  # select some at certain quantiles (best, worst, 25, 50 & 75% of dist.):
  slice(c(1, round(n() * c(0.25, 0.50, 0.75)), n()))
```

Graphically:

```{r}

selected |> 
  mutate(
    percentile_label = paste0(round(percentile), "th percentile"),
    label = paste(
      "Gender:", gender, "\n",
      "Age:", age, "\n",
      "Race:", race, "\n",
      "Disability:", disability_status, "\n",
      "Nationality:", nationality, "\n",
      "Education:", education, "\n",
      "Studies:", field_of_study, "\n",
      "Cover letter:", cover_letter, "\n",
      "Recommendations:", recommendations, "\n",
      "Experience:", experience
    )
  ) |> 
  ggplot(aes(x = estimate, y = forcats::fct_reorder(label, estimate))) +
  geom_vline(xintercept = omean, lty = "dotted") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  geom_text(
    aes(x = 1.1, label = percentile_label), 
    hjust = 1, 
    size = 3, 
    color = "grey50"
  ) +
  labs(
    x = "Estimated Pr(Hired) (LPM)",
    y = NULL,
    caption = "Vertical line = marginal hiring probability"
  )
```

