---
title: "Comparing five different scenarios on Qwen2-7B-Instruct"
editor: source
date: today
format:
  html:
    embed-resources: true
    message: false
    warning: false
    highlight-style: pygments
editor_options:
  chunk_output_type: console
---

# Packages

```{r}
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)

source(here::here("code/R/tools.r"))

theme_set(theme_nice())
```

# Data

```{r}
PROC <- here::here("code/modular/outputs/processed")

# Shared factor levels across all scenarios
set_shared_levels <- function(df) {
  df |>
    mutate(
      gender = factor(gender, levels = c("Male", "Female", "Non-binary")),
      age    = factor(age,    levels = c("30", "35", "40", "45", "50", "55")),
      race   = factor(race,   levels = c("White", "Black", "Hispanic", "Asian")),
      disability_status = factor(
        disability_status,
        levels = c("None stated", "Physical disability", "Mental health history")
      ),
      nationality = factor(
        nationality,
        levels = c("American", "Canadian", "German", "Indian", "Mexican")
      )
    )
}

financial_analyst <- readr::read_csv(
  file.path(PROC, "L_Qwen_Financial_Analyst_PositiveFrame_long.csv")
) |>
  set_shared_levels() |>
  mutate(
    education      = factor(education, levels = c("Associate Degree", "Bachelor Degree", "Master Degree")),
    field_of_study = factor(field_of_study, levels = c("Unrelated", "Somewhat related", "Closely related")),
    cover_letter   = factor(cover_letter, levels = c("Unpersonalized", "Personalized")),
    recommendations = factor(recommendations, levels = c("Below average", "Average", "Above average")),
    experience     = factor(experience, levels = c("0-2 years", "3-7 years", "10+ years"))
  )

registered_nurse <- readr::read_csv(
  file.path(PROC, "L_Qwen_Registered_Nurse_long.csv")
) |>
  set_shared_levels() |>
  mutate(
    education      = factor(education, levels = c("Associate Degree", "Bachelor Degree", "Master Degree")),
    field_of_study = factor(field_of_study, levels = c("Unrelated", "Somewhat related", "Closely related")),
    cover_letter   = factor(cover_letter, levels = c("Unpersonalized", "Personalized")),
    recommendations = factor(recommendations, levels = c("Below average", "Average", "Above average")),
    experience     = factor(experience, levels = c("0-2 years", "3-7 years", "10+ years"))
  )

software_engineer <- readr::read_csv(
  file.path(PROC, "L_Qwen_Software_Engineer_long.csv")
) |>
  set_shared_levels() |>
  mutate(
    education      = factor(education, levels = c("Associate Degree", "Bachelor Degree", "Master Degree")),
    field_of_study = factor(field_of_study, levels = c("Unrelated", "Somewhat related", "Closely related")),
    cover_letter   = factor(cover_letter, levels = c("Unpersonalized", "Personalized")),
    recommendations = factor(recommendations, levels = c("Below average", "Average", "Above average")),
    experience     = factor(experience, levels = c("0-2 years", "3-7 years", "10+ years"))
  )

personal_loan <- readr::read_csv(
  file.path(PROC, "L_Qwen_Personal_Loan_long.csv")
) |>
  set_shared_levels() |>
  mutate(
    annual_income     = factor(annual_income, levels = c("$35,000", "$60,000", "$95,000", "$140,000")),
    credit_score      = factor(credit_score, levels = c("Fair", "Good", "Excellent")),
    employment_length = factor(employment_length, levels = c("Less than 1 year", "2-5 years", "6+ years")),
    existing_debt     = factor(existing_debt, levels = c("High", "Moderate", "Low")),
    requested_amount  = factor(requested_amount, levels = c("$30,000", "$15,000", "$5,000"))
  )

rental_tenancy <- readr::read_csv(
  file.path(PROC, "L_Qwen_Rental_Tenancy_long.csv")
) |>
  set_shared_levels() |>
  mutate(
    monthly_income    = factor(monthly_income, levels = c("2x the rent", "3x the rent", "4x the rent")),
    rental_history    = factor(rental_history, levels = c("Prior late payments", "No prior rental history", "Positive references")),
    employment_status = factor(employment_status, levels = c("Part-time", "Self-employed", "Full-time")),
    credit_check      = factor(credit_check, levels = c("Fair", "Good", "Excellent")),
    deposit_offered   = factor(deposit_offered, levels = c("1 month deposit", "2 months deposit", "3 months deposit"))
  )
```

# AMCEs per scenario

```{r}
amce_financial_analyst <- amce(
  chosen ~ gender + age + race + disability_status + nationality +
    education + field_of_study + cover_letter + recommendations + experience,
  data = financial_analyst
)

amce_registered_nurse <- amce(
  chosen ~ gender + age + race + disability_status + nationality +
    education + field_of_study + cover_letter + recommendations + experience,
  data = registered_nurse
)

amce_software_engineer <- amce(
  chosen ~ gender + age + race + disability_status + nationality +
    education + field_of_study + cover_letter + recommendations + experience,
  data = software_engineer
)

amce_personal_loan <- amce(
  chosen ~ gender + age + race + disability_status + nationality +
    annual_income + credit_score + employment_length + existing_debt + requested_amount,
  data = personal_loan
)

amce_rental_tenancy <- amce(
  chosen ~ gender + age + race + disability_status + nationality +
    monthly_income + rental_history + employment_status + credit_check + deposit_offered,
  data = rental_tenancy
)
```

## Financial Analyst

```{r}
amce_financial_analyst |>
  mutate(
    attr  = attr |> gsub("_", " ", x = _) |> stringr::str_to_sentence(),
    label = stars(p.value)
  ) |>
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_text(aes(label = label), position = position_nudge(y = .15), show.legend = FALSE) +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    title = "Financial Analyst",
    x = "AMCE", y = "",
    caption = "*** p<0.01  ** p<0.05  * p<0.1"
  )
```

## Registered Nurse

```{r}
amce_registered_nurse |>
  mutate(
    attr  = attr |> gsub("_", " ", x = _) |> stringr::str_to_sentence(),
    label = stars(p.value)
  ) |>
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_text(aes(label = label), position = position_nudge(y = .15), show.legend = FALSE) +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    title = "Registered Nurse",
    x = "AMCE", y = "",
    caption = "*** p<0.01  ** p<0.05  * p<0.1"
  )
```

## Software Engineer

```{r}
amce_software_engineer |>
  mutate(
    attr  = attr |> gsub("_", " ", x = _) |> stringr::str_to_sentence(),
    label = stars(p.value)
  ) |>
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_text(aes(label = label), position = position_nudge(y = .15), show.legend = FALSE) +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    title = "Software Engineer",
    x = "AMCE", y = "",
    caption = "*** p<0.01  ** p<0.05  * p<0.1"
  )
```

## Personal Loan

```{r}
amce_personal_loan |>
  mutate(
    attr  = attr |> gsub("_", " ", x = _) |> stringr::str_to_sentence(),
    label = stars(p.value)
  ) |>
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_text(aes(label = label), position = position_nudge(y = .15), show.legend = FALSE) +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    title = "Personal Loan",
    x = "AMCE", y = "",
    caption = "*** p<0.01  ** p<0.05  * p<0.1"
  )
```

## Rental Tenancy

```{r}
amce_rental_tenancy |>
  mutate(
    attr  = attr |> gsub("_", " ", x = _) |> stringr::str_to_sentence(),
    label = stars(p.value)
  ) |>
  ggplot(aes(x = estimate, y = level, color = attr)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point() +
  geom_text(aes(label = label), position = position_nudge(y = .15), show.legend = FALSE) +
  geom_linerange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~attr, scales = "free_y", ncol = 2) +
  theme(legend.position = "none") +
  labs(
    title = "Rental Tenancy",
    x = "AMCE", y = "",
    caption = "*** p<0.01  ** p<0.05  * p<0.1"
  )
```

# Cross-scenario comparison: sensitive attributes

Stack all AMCEs and filter to the shared sensitive attributes to compare bias directly across scenarios.

```{r}
sensitive_attrs <- c("gender", "age", "race", "disability_status", "nationality")

all_amce <-
  list(
    "Financial Analyst" = amce_financial_analyst,
    "Registered Nurse"  = amce_registered_nurse,
    "Software Engineer" = amce_software_engineer,
    "Personal Loan"     = amce_personal_loan,
    "Rental Tenancy"    = amce_rental_tenancy
  ) |>
  bind_rows(.id = "scenario") |>
  filter(attr %in% sensitive_attrs) |>
  mutate(
    scenario = factor(scenario, levels = c(
      "Financial Analyst", "Registered Nurse", "Software Engineer",
      "Personal Loan", "Rental Tenancy"
    )),
    attr = attr |> gsub("_", " ", x = _) |> stringr::str_to_sentence(),
    label = stars(p.value)
  )
```

## By sensitive attribute (scenarios as color)

```{r}
#| fig-height: 10
all_amce |>
  ggplot(aes(x = estimate, y = level, color = scenario, shape = scenario)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point(position = position_dodge(.5)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.5)
  ) +
  facet_wrap(~attr, scales = "free_y", ncol = 1) +
  labs(
    title = "Sensitive attribute AMCEs across scenarios (Qwen)",
    x = "AMCE",
    y = "",
    color = "Scenario",
    shape = "Scenario",
    caption = "Estimates relative to baseline level of each attribute"
  ) +
  theme(legend.position = "bottom")
```

## By scenario (sensitive attributes as rows)

```{r}
#| fig-height: 10
all_amce |>
  ggplot(aes(x = estimate, y = level, color = attr, shape = attr)) +
  geom_vline(xintercept = 0, lty = "dotted") +
  geom_point(position = position_dodge(.4)) +
  geom_linerange(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(.4)
  ) +
  facet_wrap(~scenario, ncol = 1) +
  labs(
    title = "Sensitive attribute AMCEs by scenario (Qwen)",
    x = "AMCE",
    y = "",
    color = "Attribute",
    shape = "Attribute",
    caption = "Estimates relative to baseline level of each attribute"
  ) +
  theme(legend.position = "bottom")


#write.csv(
#  all_amce,
#  here::here("code", "modular", "amce_allscenarios.csv"),
#  row.names = FALSE
#)
```
